---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# educationdata

Retrieve data from the Urban Institute's [Education Data API](https://ed-data-portal.urban.org/) as a `data.frame` for easy analysis.

## Installation

To install the alpha version of the package:

* Clone the repository from GitHub.

`git clone https://github.com/UI-Research/education-data-package-r.git`

* Open the `educationdata.Rproj` and build from source in the R Console. Note 
that you will need to have the `devtools` package installed.

```{r gh-installation, eval=FALSE}
# install.packages(devtools) # if necessary
devtools.install()
```

## Usage

Use the `get_education_data()` function to retrieve a `data.frame` from an API 
endpoint. 

```{r usage, eval=FALSE}
get_education_data(level, source, topic, ..., by, filters)
```

The `get_education_data()` function will parse and validate several arguments 
against the API.

* level - API data level to query. Current levels are:
    - college-university
    - school-districts
    - schools
* source - API data source to query. Current sources are:
    - ccd
    - ipeds
    - saipe
* topic - API data topic to query
* ... - Additional parameters required for a specific API call
* by - Additional optional parameters for a specific API call
* filters - Optional query to filter the resutls from an API call

## Available Endpoints

```{r endpoints, echo=FALSE}
df <- educationdata::get_endpoint_info()
vars <- c('endpoint_url', 
          'years_available')
df <- df[vars]
df <- df[order(df$endpoint_url), ]
knitr::kable(df[vars], 
             col.names = c('Endpoint', 'Years Available'),
             row.names = FALSE)
```


## Examples

Let's build up some examples, starting with a simple API endpoint:

* `/api/v1/schools/ccd/enrollment/{year}/{grade}/`

In this case, the `level` is `schools`, the `source` is `ccd`, and the `topic` is 
`enrollment`. Note that variables in curly brackets `year` and `grade` are 
required for this endpoint.

```{r example-01, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8')
```

Note that this endpoint is also callable `by` certain variables:

* `/api/v1/schools/ccd/enrollment/{year}/{grade}/race/`
* `/api/v1/schools/ccd/enrollment/{year}/{grade}/sex/`
* `/api/v1/schools/ccd/enrollment/{year}/{grade}/race/sex/`

These optional variables can be added to the `by` argument:

```{r example-02, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8',
                         by = list('race', 'sex'))
```

You may also filter the results of an API call:

```{r example-03, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8',
                         by = list('race', 'sex'),
                         filters = list('ncessch' = '010000200277'))
```
