---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.path = 'README-'
)
```

# educationdata

Retrieve data from the Urban Institute's [Education Data API](https://ed-data-portal.urban.org/) as a `data.frame` for easy analysis.

## Installation

To install the alpha version of the package:

* Clone the repository from GitHub.

`git clone https://github.com/UI-Research/education-data-package-r.git`

* If you have RStudio installed, open `educationdata.Rproj` and build from 
source in the console. If you are using the R GUI, set your working directory 
to the project folder. 

* Install the `devtools` package if you don't already have it, and run:

```{r gh-installation, eval=FALSE}
# install.packages(devtools) # if necessary
devtools::install()
```

## Endpoint Format

API endpoints follow a format of

  * `level/source/topic/{required}/{variables}/grouping/variables`

where:

  * `level` is the data level to query (i.e., schools, school-districts, 
  college-university)
  * `source` is the data source for a specific level (i.e., ccd, ipeds, saipe)
  * `topic` is the data topic to query
  * `{required}/{variables}` are arguments that must be set for a specific topic
  * `grouping/variables` are arguments that will provide data 'by' certain 
  categories
  
For example, the endpoint 

  * `college-university/ipeds/fall-enrollment/{year}/{level_of_study}/`

will provide:

  * college-university level data 
  * from IPEDS 
  * on fall-enrollment numbers 
  * for a specified year and level of study.

While the endpoint 

  * `college-university/ipeds/fall-enrollment/{year}/{level_of_study}/race`
  
will provide the same data, grouped by race.

## Available Endpoints

```{r endpoints, echo=FALSE}
url <- 'https://ed-data-portal.urban.org/api/v1/api-endpoints/'
res <- httr::GET(url)
df <- jsonlite::fromJSON(rawToChar(res$content))$results

vars <- c('endpoint_url', 
          'years_available')
df <- df[vars]

df$endpoint_url <- gsub('/api/v1/', '', df$endpoint_url)
df$years_available <- gsub('–' ,'-', df$years_available)
df <- df[order(df$endpoint_url), ]

knitr::kable(df[vars], 
             col.names = c('Endpoint', 'Years Available'),
             row.names = FALSE)
```

## Usage

Use the `get_education_data()` function to retrieve a `data.frame` from an API 
endpoint. 

```{r usage, eval=FALSE}
get_education_data(level, source, topic, ..., by, filters)
```

The `get_education_data()` function will parse and validate several arguments 
against the API.

* level - API data level to query. Current levels are:
    - college-university
    - school-districts
    - schools
* source - API data source to query. Current sources are:
    - ccd
    - ipeds
    - saipe
* topic - API data topic to query
* ... - Additional parameters required for a specific API call
* by - Additional optional parameters for a specific API call
* filters - Optional query to filter the results from an API call


## Examples

Let's build up some examples, starting with a simple API endpoint:

* `schools/ccd/enrollment/{year}/{grade}/`

In this case, the `level` is `schools`, the `source` is `ccd`, and the `topic` is 
`enrollment`. Note that variables in curly brackets (i.e., `year` and `grade`) 
are required for this endpoint.

```{r example-01, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8')
```

Note that this endpoint is also callable `by` certain variables:

* `schools/ccd/enrollment/{year}/{grade}/race/`
* `schools/ccd/enrollment/{year}/{grade}/sex/`
* `schools/ccd/enrollment/{year}/{grade}/race/sex/`

These optional variables can be added to the `by` argument:

```{r example-02, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8',
                         by = list('race', 'sex'))
```

You may also filter the results of an API call:

```{r example-03, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8',
                         by = list('race', 'sex'),
                         filters = list('ncessch' = '010000200277'))
```
