---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = '#>',
  fig.path = 'README-'
)
```

# educationdata

[![Travis-CI Build Status](https://travis-ci.org/UrbanInstitute/education-data-package-r.svg?branch=master)](https://travis-ci.org/UrbanInstitute/education-data-package-r)

Retrieve data from the Urban Institute's [Education Data API](https://ed-data-portal.urban.org/) as a `data.frame` for easy analysis.

**Note:** This package is still currently under **active** development. Please 
file all bugs, issues, and suggestions as an 
[Issue in the GitHub repository](https://github.com/UrbanInstitute/education-data-package-r/issues) 
for the development team. 

## Installation

To install the package:

* Install the `devtools` package if you don't already have it, and run:

```{r gh-installation, eval=FALSE}
# install.packages('devtools') # if necessary
devtools::install_github('UrbanInstitute/education-data-package-r')
```

## Quick Start Guide

```{r quickstart-01, message=FALSE}
library(educationdata)

df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2014,
                         grade = 'grade-12',
                         by = list('race', 'sex'),
                         filters = list('ncessch' = '340606000122'),
                         add_labels = TRUE)

head(df)
```

The `get_education_data()` function will return a `data.frame` from a call to 
the Education Data API.  

```{r quickstart, eval=FALSE}
get_education_data(level, source, topic , ..., by, filters, add_labels)
```

where:

* level (required) - API data level to query.
* source (requried) - API data source to query.
* topic (required) - API data topic to query.
* ... (required) - Additional parameters required for a specific endpoint.
* by (optional) - Optional parameters for a specific API call.
* filters (optional) - Optional query to filter the results from an API call.
* add_labels - Add variable labels (when applicable)? Defaults to `FALSE`.

## Endpoint Format

API endpoints follow a format of

  * `level/source/topic/{required}/{variables}/grouping/variables`

where:

  * `level` is the data level to query (i.e., schools, school-districts, 
  college-university)
  * `source` is the data source for a specific level (i.e., ccd, ipeds, saipe)
  * `topic` is the data topic to query
  * `{required}/{variables}` are arguments that must be set for a specific topic
  * `grouping/variables` are arguments that will provide data 'by' certain 
  categories
  
For example, the endpoint 

  * `college-university/ipeds/fall-enrollment/{year}/{level_of_study}/`

will provide:

  * college-university level data 
  * from IPEDS 
  * on fall-enrollment numbers 
  * for a specified year and level of study.

While the endpoint 

  * `college-university/ipeds/fall-enrollment/{year}/{level_of_study}/race`
  
will provide the same data, grouped by race.

## Available Endpoints

```{r endpoints, echo=FALSE}
url <- 'https://ed-data-portal.urban.org/api/v1/api-endpoints/'
res <- httr::GET(url)
df <- jsonlite::fromJSON(rawToChar(res$content))$results

vars <- c('endpoint_url', 
          'years_available')
df <- df[vars]

df$endpoint_url <- gsub('/api/v1/', '', df$endpoint_url)
df$years_available <- gsub('and' ,'', df$years_available)
df$years_available <- gsub('\u20AC' ,'-', df$years_available)
df$years_available <- gsub('\u00E2' ,'', df$years_available)
df$years_available <- gsub('\u201C' ,'', df$years_available)
df <- df[order(df$endpoint_url), ]

knitr::kable(df[vars], 
             col.names = c('Endpoint', 'Years Available'),
             row.names = FALSE)
```

## Required Variables

In addition to `year`, the other requried variables for certain endpoints 
accept the following values:

### Grade
| Function Argument | Grade |
|-------------------|-------|
| `grade = 'grade-pk'` | Pre-K  |
| `grade = 'grade-k'`  | Kindergarten  |
| `grade = 'grade-1'` | Grade 1  |
| `grade = 'grade-2'` | Grade 2  |
| `grade = 'grade-3'` | Grade 3  |
| `grade = 'grade-4'` | Grade 4  |
| `grade = 'grade-5'` | Grade 5  |
| `grade = 'grade-6'` | Grade 6  |
| `grade = 'grade-7'` | Grade 7  |
| `grade = 'grade-8'` | Grade 8  |
| `grade = 'grade-9'` | Grade 9  |
| `grade = 'grade-10'` | Grade 10  |
| `grade = 'grade-11'` | Grade 11  |
| `grade = 'grade-12'` | Grade 12  |
| `grade = 'grade-13'` | Grade 13  |
| `grade = 'grade-14'` | Adult Education |
| `grade = 'grade-15'` | Ungraded  |
| `grade = 'grade-16'` | K-12  |
| `grade = 'grade-20'` | Grades 7 and 8  |
| `grade = 'grade-21'` | Grade 9 and 10  |
| `grade = 'grade-22'` | Grades 11 and 12  |
| `grade = 'grade-99'` | Total  |

### Level of Study
| Function Argument | Level of Study |
|-------------------|----------------| 
| `level_of_study = 'undergraduate'` | Undergraduate |
| `level_of_study = 'graduate'` | Graduate |
| `level_of_study = 'first-professional'` | First Professional |
| `level_of_study = 'post-baccalaureate'` | Post-baccalaureate |
| `level_of_study = '99'` | Total |


## Usage

Use the `get_education_data()` function to retrieve a `data.frame` from an API 
endpoint. 

```{r usage, eval=FALSE}
get_education_data(level, source, topic, ..., by, filters, add_labels)
```

The `get_education_data()` function will parse and validate several arguments 
against the API.

* level (required) - API data level to query. Current levels are:
    - college-university
    - school-districts
    - schools
* source (required) - API data source to query. Current sources are:
    - ccd
    - ipeds
    - saipe
* topic (required) - API data topic to query.
* ... (required) - Additional parameters required for a specific API call.
* by - Optional parameters for a specific API call.
* filters - Optional query to filter the results from an API call.
* add_labels - Add variable labels (when applicable)? Defaults to `FALSE`.


## Examples

Let's build up some examples, starting with a simple API endpoint:

* `schools/ccd/enrollment/{year}/{grade}/`

In this case, the `level` is `schools`, the `source` is `ccd`, and the `topic` is 
`enrollment`. Note that variables in curly brackets (i.e., `year` and `grade`) 
are required for this endpoint.

```{r example-01-a, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8')
```

The `year` parameter is vectorized and can accept multiple years:

```{r example-01-b, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004:2008,
                         grade = 'grade-8')
```

Or the endpoint can be called for 'all' `years`:

```{r example-01-c, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 'all',
                         grade = 'grade-8')
```

Note that this endpoint is also callable `by` certain variables:

* `schools/ccd/enrollment/{year}/{grade}/race/`
* `schools/ccd/enrollment/{year}/{grade}/sex/`
* `schools/ccd/enrollment/{year}/{grade}/race/sex/`

These optional variables can be added to the `by` argument:

```{r example-02, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8',
                         by = list('race', 'sex'))
```

You may also filter the results of an API call:

```{r example-03, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2004,
                         grade = 'grade-8',
                         by = list('race', 'sex'),
                         filters = list('ncessch' = '010000200277'))
```

Finally, the `add_labels` flag will map variables to a `factor` from their 
labels in the API.

```{r example-04, eval=FALSE}
df <- get_education_data(level = 'schools', 
                         source = 'ccd', 
                         topic = 'enrollment', 
                         year = 2014,
                         grade = 'grade-8',
                         add_labels = TRUE)
```
